Self Referencing Sequences

### 2014-11-01 draft.tex
edited up to iterative process, removed prefixes

\section{Prefixes}
\label{prefixes}
\emph{Written by Perry Kleinhenz, edited by Fermi Ma and Erik Waingarten}
\begin{definition} We say that two sequences $\{a_i\}$ and $\{b_i\}$ differ by a prefix of length $n$ if 
\begin{equation*}
a_{n+i} = b_{i},
\end{equation*}
for $i$ any positive integer. We say that two sequences are independent if they do not differ by a prefix for any $n$, in either order. 
\end{definition}
Note that a pair of sequences can differ by multiple prefixes of different length. For example if we have 
\begin{align*}
&\{a_i\} = \{1,2,3,4,5,6,1,2,3,4,5,6, \ldots \} \\  
&\{b_i\}= \{7,8,9,1,2,3,4,5,6,1,2,3,4,5,6, \ldots\},
\end{align*}
 then $a_i$ and $b_i$ differ by prefixes of length $3+6k$ for $k$ a nonnegative integer. Because of this we make the following definition.
 \begin{definition} 
 We say that two sequences $\{a_i\}$ and $\{b_i\}$ differ by a minimal prefix of length $n$ if they differ by a prefix of length $n$, but do not differ by a prefix of length $m$ for all $0<m<n$.
 \end{definition}
We note that if a pair of sequences differs by a prefix they differ by a minimal prefix. In our above example $a_i$ and $b_i$ differ by a minimal prefix of length $3$. The following lemma is a direct consequence of our algorithm for generating self-referential sequences:

\begin{lemma} Let $G = \{1, g_1, g_2, \ldots, g_n\}$, $s(G, 1)$ and $s(G, g_1)$ differ by a minimal prefix of length at most 1.
\end{lemma}
\begin{proof}
Let $a_i$ refer to the $i$th term of $s(G, 1)$ and $b_i$ refer to the $i$th term of $s(G, g_1)$. If we use our algorithm for generating the sequence, we note that $a_1 = 1$, and so $a_2 = g_1$, and that this is the beginning of a new block of length $g_1$. Furthermore, $b_1 = g_1$ and it indicated the beginning of a block of length $g_1$. Note that our algorithm is ``memory-less'' in the sense that it does not maintain any state about previous elements. Because the number and the begining of the block is the same. The algorithm will generate the same sequence, shifted by 1. Therefore, $s(G,1)$ and $s(G, g_1)$ differ by a minimal prefix of length at most 1.
\end{proof}
We now give a 
\begin{theorem}
\label{independentsequences}
Let $G = \{ 1, a, b\}$ where $1<a<b$. $s(G, a)$ and $s(G, b)$ are independent.
\end{theorem}
\begin{proof}
We proceed by contradiction. If the two sequences are not independent then they must differ by some minimal prefix of length $n$. If we let $a_i$ be the sequence beginning with $a$ and $b_i$ be the sequence beginning with $b$ then
\begin{equation*}
a_{n+k} = b_{k}.
\end{equation*}
We know that the first $b$ terms of $b_i$ are $b$'s. Because $b>a>1$ no block can be longer then $b$. Thus the terms immediately preceding and following this block in  $a_i$ cannot be $b$'s. Because of this the first block of $b_i$ is the $m+1$th block of $a_i$, for some $m$ a positive integer. Since the value of the sequence at position $k$ is the length of the $k$th block this means that $a_{m+1}=b_1$. In fact the $k$th block of $b_i$ is the $(m+k)$th block of $a_i$ and so 
\begin{equation*}
a_{m+k} = b_k
\end{equation*}

We now claim that the prefix has length strictly larger than $m$. Assume otherwise so $n \leq m$. 
This means that $a_{m+1}$ which gives the length of the $m+1$th block occurs at or after the beginning of the $m+1$th block, which begins at position $a_{n+1}$. By our proof that self-referential sequences are well defined we know that the second case cannot occur. Therefore we must have the $m+1$th block start at position $m+1$. This would require every block to have length exactly 1, but this is a contradiction because the first block has length $a_1=a>1$. Therefore the prefix must have length strictly larger than $m$.

We note that based on a previous step the two sequences differ by a prefix starting at $m$. Since $m<n$ we have produced a prefix which starts before $n$ which is a contradiction. Therefore the two sequences are independent. 
\end{proof}

@@

In this section, we consider a slightly weaker definition of equivalence. Consider, for example, sequences generated by $G = \{1,2,3\}$. Theorem~\ref{independentsequences} shows that the sequence starting with 2
\begin{align*}
2,2,3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots
\end{align*}
and the sequence starting with 3
\begin{align*}
3,3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots
\end{align*}
are independent in the sense that neither sequence is exactly contained in the other. However, if we delete the first two numbers of the sequence starting with 2, and the first number of the sequence starting with 3, we get 
\begin{align*}
3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots
\end{align*}
in both cases. Lets look at the state of the algorithm that generates both sequences. After the first block, the execution of the algorithm of $s(G, 2)$ right after placing the first $3$'s (in the 3rd and 4th position), the algorithm is in state corresponding to:
\begin{itemize}
\item block size $3$
\item no elements in the current block
\item starting number is $1$
\end{itemize}.   
If we look at the execution of the algorithm that generates $s(G, 3)$ after placing the first three $3$'s, the algorithm is in state corresponding to:
\begin{itemize}
\item block size $3$
\item no elements in the current block
\item starting number is $1$
\end{itemize}
Therefore, since the states are the same, the sequence generated after the 4th position of $s(G, 2)$ and after the $3rd$ position of $s(G, 3)$ will be the same. By inspection, we see we can also include two of the 3's.

Thus, since both sequences are essentially the same if a prefix is removed from each of them, we call these sequences \emph{equivalent}. To formalize, this, we say that two sequences $\{a_i\}$ and $\{b_i\}$ are equivalent if there exists some integers $n,k \geq 0$ such that
\begin{align*}
a_{n+i} = b_{k+i}
\end{align*}

We now give the following conjecture

@@

This conjecture is partly motivated by the fact that the property holds for $\{1,2,3\}$. For generating sets with larger numbers, we have been unable to verify or refute the conjecture. However, in Section~\ref{density}, we show that for some generating sets the starting number of a sequence does not affect the densities of each number, which provides some evidence for this conjecture.

We now introduce a notion which will help us categorize sequences as equivalent. 
\begin{definition} Consider a sequence $a$. We define the read distance for an element $a_i$ as the number of terms in between it and the beginning of the $i$th block of numbers.
\end{definition}
The read distance can be any nonnegative integer. For example if we consider the sequence $a_i$ generated by $\{1,2,3\}$ beginning with a 2  
\begin{align*}
2,2,3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots
\end{align*}
the read distance for $a_1$ (the first two) is $0$ as the first block of numbers starts at the first number, the read distance for $a_2$ (the second two) is $1$ as the second block of numbers starts at the third number. We can continue to calculate the read distances for this sequence and see that they form a monotonically increasing sequence $0,1,2,4,6,6,6,6,7,8,9,\ldots$. From this we might suspect that the read distance for any self referencing sequence will also be monotonically increasing. We can in fact compute the read distance exactly. 
\begin{lemma} 
\label{readdistformula}
Let $a_i$ be any self referencing sequence generated by some set of positive integers. Then the read distance for the element $a_i$, for $i>1$, is 
\begin{equation*}
\sum_{k=1}^{i-1}( a_k - 1)
\end{equation*} 
\end{lemma}
\begin{proof}
We will prove this using induction. We know that for $i=1$ the read distance is $0$, as the first block always starts at the first number. When $i=2$ we must show that the read distance is $a_1-1$. The first block will be of length $a_1$ since the sequence is self referencing, therefore the second block will begin at position $a_1$. The number of elements between position $a_1$ and $2$ is exactly $a_1 -1$.

@@

\begin{lemma} Consider a self referencing sequence $x$ generated by $\{a_1, a_2, \ldots, a_n\}$ with $a = \max(a_1, a_2, \ldots, a_n)$. Suppose the read distance for $x_i$ is $d$, then the read distance for $x_{i+a}$ is at least $d+1$. 
\end{lemma}
\begin{proof}
Assume otherwise. By Lemma ~\ref{readdistformula} we know that the read distance for $x_{i+a}$ is at least $d$. Thus it must be exactly $d$. Lemma ~\ref{readdistformula} also tells us that if this is the case then $x_{i+k}=1$ for $k=0,1,2,\ldots, a_n$. But this means that we have a block of length $a+1$, which is a contradiction of the self referencing property of the sequence. Therefore the read distance for $x_{i}$ is at least $d+1$. 
\end{proof}

@@

We also have a sufficient condition for equivalence 
\begin{lemma} If two sequences $a_i, b_i$ are such that for some integers $m,l$ the read distance for $a_m$ and $b_l$ are the same and 
\begin{equation*}
a_{m+k} = b_{l+k}
\end{equation*}
where $k$ ranges from 0 to the read distance for $a_m$ then $a_i$ and $b_i$ are equivalent. 
\end{lemma}
\begin{proof}
We will prove this using a simple induction. Suppose the read distance for $a_m$ is $d$ then know that $a_{m+d} = b_{l+d}$ because $a_{m+d-1} = b_{l+d-1}$ and $a_{m+d}$ is the beginning of a new block. Furthermore we have that the read distance for $a_{m+d}$ equals the read distance for $b_{l+d}$. Now suppose that we have for $k \leq j$ that $a_{m+k} = b_{l+k}$ and the read distance for $a_{m+k}$ equals the read distance for $b_{l+k}$ and we would like to show that this holds for $k = j+1$. By Lemma ~\ref{readdistformula} we actually already know that the read distance for the two is the same, as it is the same for the preceding terms, which are equal.

@@

Otherwise $a_{m+k}$ and $a_{m+k+1}$ are in the same block, then we only need to show that $b_{l+k}$ and $b_{l+k+1}$ are in the same block as well. But as we argued above the length of blocks is determined by terms earlier in the sequence which we know to be equal and have the same read distance. Therefore we have $a_{m+k+1} = b_{l+k+1}$ for this case as well. This completes our induction and so $a_i$ and $b_i$ are equivalent. 
\end{proof}
These conditions for equivalence lead us to make the following conjecture on a sufficient condition for nonequivalence.

\begin{conjecture} If two sequences $a_i$ and $b_i$ have elements $a_k \neq b_l$, for $k,l>1$ with the same read distance then they are not equivalent. 
\end{conjecture}

>>>>

@@

In this section, we consider a different definition of equivalence. Consider, for example, sequences generated by $C = (1,2,3)$. Theorem~\ref{independentsequences} shows that the sequence starting with 2
\begin{align*}
2,2,3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots
\end{align*}
and the sequence starting with 3
\begin{align*}
3,3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots
\end{align*}
are independent in the sense that neither sequence is exactly contained in the other. However, if we delete the first two numbers of the sequence starting with 2, and the first number of the sequence starting with 3, we get 
\begin{align*}
2,2,|3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots\\
3,|3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots
\end{align*}
in both cases.

Thus, since both sequences are essentially the same if a prefix is removed from each of them, we call these sequences \emph{equivalent}. To formalize, this, we say that two sequences $(a_i)$ and $(b_i)$ are equivalent if there exists some integers $n,k \geq 0$ such that
\begin{align*}
a_{n+i} = b_{k+i}
\end{align*}
for all positive integers $i$. This definition is saying that the sequences are eventually equivalent.

@@

This conjecture is partly motivated by the fact that the property holds for self-referencing sequences from the generating cycle $C= (1,2,3)$. For generating cycles with larger numbers, we have been unable to verify or refute the conjecture.. However, in Section~\ref{density}, we show that for some generating sets the starting number of a sequence does not affect the densities of each number, which provides some evidence for this conjecture.

We now introduce a notion which will help us categorize sequences as equivalent. 
\begin{definition} Consider a sequence $(a_i)$. We define the read distance for an index $i$ as the number of terms in between it $a_i$ and the beginning of the $i$th block of numbers.
\end{definition}
The read distance can be any nonnegative integer. For example if we consider the sequence $(a_i)$ generated by $(1,2,3)$ beginning with a 1  
\begin{align*}
1,2,2,3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots
\end{align*}
the read distance for $1$ is always $0$ since in a self-referencing sequence, the first number must refer to the first block. The read distance for $2$ is $0$ as the second block of numbers starts at the second number. We can continue to calculate the read distances for this sequence and see that they form a monotonically increasing sequence $0, 0 ,1,2,4,6,6,6,6,7,8,9,\ldots$. The read distance, can be thought of as the horizontal shift of the arrows in Figure~\ref{fig:onetwothree}.

\begin{lemma} 
\label{readdistformula}
Let $(a_i)$ be any self referencing sequence generated by some set of positive integers. Then the read distance for $i>1$, is 
\begin{equation*}
\sum_{k=1}^{i-1}( a_k - 1)
\end{equation*} 
and the read distance for $1$ is always 0.
\end{lemma}
\begin{proof}
We will prove this using induction. We know that for $i=1$ the read distance is $0$, as the first block always starts at the first number. When $i=2$ the read distance is $a_1-1$. The first block will be of length $a_1$ since the sequence is self referencing, therefore the second block will begin at position $a_1$. The number of elements between position $a_1$ and $2$ is exactly $a_1 -1$.

@@

\begin{lemma} Consider a self referencing sequence $(a_i)$ generated by $(c_1, c_2, \ldots, c_n)$ with $m = \max(c_1, c_2, \ldots, c_n)$. Suppose the read distance for $i$ is $d$, then the read distance for $i+m$ is at least $d+1$. 
\end{lemma}
\begin{proof}
The read distances can stay constant from $i$ to $i+1$ only if there is a $a_{i+1} = 1$. There can be at most $m$ 1's in a contiguous block, so the read distance of $i+m$ is bigger than the read distance of $i$.  
\end{proof}

@@

We also have a sufficient condition for equivalence 
\begin{lemma} If two sequences $(a_i), (b_i)$ are such that for some integers $m,l$ the read distance for $a_m$ and $b_l$ are the same and 
\begin{equation*}
a_{m+k} = b_{l+k}
\end{equation*}
where $k$ ranges from 0 to the read distance for $a_m$ then $(a_i)$ and $(b_i)$ are equivalent. 
\end{lemma}
\begin{proof}
We will prove this using a simple induction. Suppose the read distance for $a_m$ is $d$ then know that $a_{m+d} = b_{l+d}$ because $a_{m+d-1} = b_{l+d-1}$ and $a_{m+d}$ is the beginning of a new block. Furthermore we have that the read distance for $a_{m+d}$ equals the read distance for $b_{l+d}$. Now suppose that we have for $k \leq j$ that $a_{m+k} = b_{l+k}$ and the read distance for $a_{m+k}$ equals the read distance for $b_{l+k}$ and we would like to show that this holds for $k = j+1$. By Lemma ~\ref{readdistformula} we actually already know that the read distance for the two is the same, as it is the same for the preceding terms, which are equal.

@@

Otherwise $a_{m+k}$ and $a_{m+k+1}$ are in the same block, then we only need to show that $b_{l+k}$ and $b_{l+k+1}$ are in the same block as well. But as we argued above the length of blocks is determined by terms earlier in the sequence which we know to be equal and have the same read distance. Therefore we have $a_{m+k+1} = b_{l+k+1}$ for this case as well. This completes our induction and so $(a_i)$ and $(b_i)$ are equivalent. 
\end{proof}

The following is a simple observation of sequences that begin with $1$.

\begin{corollary} Let $C = (1, c_1, c_2, \ldots, c_n)$. $s(C, 1)$ and $s(C, c_1)$ are equivalent.
\end{corollary}
\begin{proof}
Let $(a_i) = s(C, 1)$ and $(b_i) = s(C, c_1)$. The read distance of sequence $(a_i)$ for $i=2$ is 0, and the read distance of sequence $(b_i)$ for $i = 1$ is 0. In addition, $a_2 = b_1 = c_1$, which means the two sequences are equivalent.
\end{proof}

These conditions for equivalence lead us to make the following conjecture on a sufficient condition for nonequivalence.

\begin{conjecture} If two sequences $(a_i)$ and $(b_i)$ have elements $a_k \neq b_l$, for $k,l>1$ with the same read distance then they are not equivalent. 
\end{conjecture}
